---
title: "Review"
author: "Lucy D'Agostino McGowan"
footer:  "[Dr. Lucy D'Agostino McGowan](https://lucymcgowan.com)"
logo: "img/icon.png"
format: 
  kakashi-revealjs: 
    theme: [custom.scss]
    slide-number: true
    chalkboard: true
    title-slide-attributes: 
      data-background-color: "#fff" 
      data-color: "#70001A"
    menu: true
---

# Make sure to bring a calculator to the exam

```{r child = "setup.Rmd"}
```

## Exam {.small}

-   Two parts
    -   Part 1: In class Monday (you can have one cheat sheet)
    -   Part 2: The same exam, taken at home -- this is open notes (there will be no class Wednesday so you can have dedicated time to work on this then)

## 

| y   | x1  | x2  |
|-----|-----|-----|
| 5.7   | 2   |  1  |
| 8.3   | 3   |  1  |
| 7.3   | 4   | 0   |

You want to predict `y` using `x1` and `x2` write out how you would calculate $\hat\beta$ in matrix form using the data provided (you do not need to solve the matrix)

## 

| y   | x1  | x2  |
|-----|-----|-----|
| 5.1   | 2   |  1  |
| 6.9   | 3   |  1  |
| 7.8   | 4   | 0   |

Solving the above equation results in the following:

$$
\begin{bmatrix}
\hat\beta_0 \\\hat\beta_1\\\hat\beta_2
\end{bmatrix} = 
\begin{bmatrix}
0.7\\1.8\\0.9
\end{bmatrix}
$$

Using the information provided, calculate the MSE for this model.

## 

| y   | x1  | x2  |
|-----|-----|-----|
| 9.1   | 4   |  1  |
| 6.2   | 3   |  0  |
| 5.8   | 2   | 1   |

You get a new test data set (above). Using the model you fit to the training data, calculate the MSE in this test set.

## Logistic Regression

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(broom)
glm(am ~ wt, data = mtcars, family = "binomial") %>%
  tidy(conf.int = TRUE) %>%
  knitr::kable(digits = 2)
```
We are predicting whether a car has automatic transmission based on it's weight

How do you interpret $\hat\beta_1$ is this a marginal or conditional effect?

## Logistic Regression

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(broom)
glm(am ~ wt + mpg, data = mtcars, family = "binomial") %>%
  tidy(conf.int = TRUE) %>%
  knitr::kable(digits = 2)
```

We add in an additional variable

How do you interpret $\hat\beta_1$? Is this a conditional effect?

## Penalized Regression {.small}

* What is the penalty for Ridge Regression?
* What is the penalty for Lasso?
* What is the equation for Elastic Net?
* How do we choose $\lambda$? $\alpha$?

## Bias-variance trade off

* What is the bias-variance trade-off?
* As the flexibility of the model increases, how does that impact bias? variance? Training MSE? Testing MSE?
* As $\lambda$ increases in penalyzed regression, how does this impact the **flexibility** of the model?

## k-fold cross validation

* How does it work?
* What are the advantages/disadvantages of a small $k$ vs large $k$